---
title: "Basic EDA plots"
output:
  html_notebook:
    df_print: paged
---

```{r}
library(RSQLite)
library(tidyverse)
library(hrbrthemes)
library(kableExtra)
library(RColorBrewer)
library(viridis)
library(d3heatmap)
library(naivebayes)
library(yardstick)
# library(ggridges)


con <- dbConnect(RSQLite::SQLite(), "../wildfire/FPA_FOD_20170508.sqlite")
causes <- dbGetQuery(con, "SELECT COUNT(*) AS nr, STAT_CAUSE_DESCR AS cause_name FROM Fires GROUP BY STAT_CAUSE_DESCR")

#knitr::opts_chunk$set(fig.width=16)

# https://www.reddit.com/r/RStudio/comments/euew3l/preview_button_gone_can_only_knit/

# some color ideas from https://colorbrewer2.org
```

```{r}
# https://www.data-to-viz.com/graph/lollipop.html

causes %>%
  arrange(nr) %>%
  mutate(cause_name=factor(cause_name, cause_name)) %>%
  ggplot( aes(x=cause_name, y=nr) ) +
    geom_segment(aes(x=cause_name, xend=cause_name, y=0, yend=nr), color="darkgrey") +
    geom_point(size=4, color='#e34a33', alpha=0.9) +
    coord_flip() +
    theme_ipsum() +
    #xlab("") +
    labs(x="",y="", title = 'Main causes for wildfires') +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      #legend.position="none",
      #axis.text = element_text(hjust = 0.5)
    )
```

```{r}
fire_size_class <- dbGetQuery(con, "SELECT COUNT(*) AS nr, FIRE_SIZE_CLASS AS fire_size FROM Fires GROUP BY FIRE_SIZE_CLASS")
```

```{r}
# https://www.data-to-viz.com/graph/barplot.html

fire_size_class %>%
  arrange(nr) %>%
  mutate(fire_size=factor(fire_size, fire_size)) %>%
  ggplot( aes(x=fire_size, y=nr) ) +
    #geom_segment(aes(x=cause_name, xend=fire_size, y=0, yend=nr), color="darkgrey") +
    #geom_point(size=4, color='#e34a33', alpha=0.9) +
    geom_bar(stat='identity', fill='darkred', width = 0.75) +
    coord_flip() +
    #theme_ipsum() +
    #xlab("") +
    labs(x="",y="", title = 'Number of fires by size') +
    theme(
      panel.grid.minor.x = element_line(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      #axis.ticks.x = fire_size_class$nr
      #legend.position="none",
      #axis.text = element_text(hjust = 0.5)
    ) # + scale_y_discrete(fire_size_class$nr)
```

What are the most frequent causes for the biggest fires?

Note that these are normalized by column (fire class).

```{r}
# https://www.data-to-viz.com/graph/heatmap.html
# https://www.r-graph-gallery.com/215-the-heatmap-function.html

fire_cause <- dbGetQuery(con,
                         "SELECT STAT_CAUSE_DESCR AS cause, FIRE_SIZE_CLASS AS fire_class, COUNT(FIRE_SIZE_CLASS) AS nr
                         FROM Fires
                         WHERE FIRE_SIZE_CLASS IN ('D', 'E', 'F', 'G')
                         GROUP BY 1, 2")


nrows <- length(levels(factor(fire_cause$cause)))
ncols <- length(levels(factor(fire_cause$fire_class)))
matr3 <- t(matrix(fire_cause$nr, nrow = ncols, ncol = nrows))
rownames(matr3) <- levels(factor(fire_cause$cause))
colnames(matr3) <- levels(factor(fire_cause$fire_class))
coul <- colorRampPalette(brewer.pal(8, "OrRd"))(25)
d3heatmap(matr3, scale = 'column', Colv = NA, Rowv = NA, col=coul,  yaxis_width = 200, width = 700, title = 'The most frequent causes for the biggest fires')
```

Debugging purposes. Previous version of this plot had wrong data.

```{r}
matr3
```
```{r}
fire_cause %>% filter(cause == 'Debris Burning')
```



```{r}
# sources: same as previous heatmap

sfire_cause <- dbGetQuery(con,
                         "SELECT STAT_CAUSE_DESCR AS cause, FIRE_SIZE_CLASS AS fire_class, COUNT(FIRE_SIZE_CLASS) AS nr
                         FROM Fires
                         WHERE FIRE_SIZE_CLASS IN ('A', 'B', 'C')
                         GROUP BY 1, 2")


snrows <- length(levels(factor(sfire_cause$cause)))
sncols <- length(levels(factor(sfire_cause$fire_class)))
smatr3 <- t(matrix(sfire_cause$nr, nrow = sncols, ncol = snrows))
rownames(smatr3) <- levels(factor(sfire_cause$cause))
colnames(smatr3) <- levels(factor(sfire_cause$fire_class))
scoul <- colorRampPalette(brewer.pal(8, "OrRd"))(25)
d3heatmap(smatr3, scale = 'column', Colv = NA, Rowv = NA, col=scoul, yaxis_width = 200, width = 700, main = 'The most frequent causes for small fires')
```


How many wildfires are managed by each unit type? The following pattern might uncover organizational issues.

```{r}
# source: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

# end of multiplot
# source: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
##############################################################################





units_by_type <- dbGetQuery(con, "SELECT count(distinct(units.UnitId)) AS nr_units, count(distinct(f.fod_id)) AS nr_fires, units.UnitType FROM Fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId GROUP BY units.UnitType")

# units_by_type
# nr_units | nr_fires | UnitType
#     7      	   40    	Interagency		
#    11        3725   	Tribe		
#   526       86871   	US County/Local		
#   902   	 477787 	  US Federal		
#   177	    1277285   	US State

# https://www.r-graph-gallery.com/37-barplot-with-number-of-observation.html
data <- data.frame(
  counts = units_by_type[,1:2],#) %>% map_int(1L)),
  units = rep(units_by_type[,3], times=1)
)

# same as the barplot above
# colors picked to show fires with some sort of red and units with some sort of blue (obvious reasons)

p1 <- data  %>% 
  #mutate(year=as.factor(year)) %>%
  ggplot( aes(x=units, y=counts.nr_fires)) +
    geom_bar(stat="identity", fill='#e6550d') +
    scale_fill_viridis(discrete=TRUE, name="") +
    theme_ipsum() +
    xlab("")+
    ylab("Number of fires")

p2 <- data  %>% 
  #mutate(year=as.factor(year)) %>%
  ggplot( aes(x=units, y=counts.nr_units)) +
    geom_bar(stat="identity", fill='#2b8cbe') +
    scale_fill_viridis(discrete=TRUE, name="") +
    theme_ipsum() +
    xlab("")+
    ylab("Number of units")
multiplot(p1, p2)

```


Further down the line, we notice many agencies each with very few units.

```{r}
# units per agency
# units names taken from kaggle dataset description

agencies <- dbGetQuery(
  con, "
    SELECT
      COUNT(*) AS nr,
      agency,
      CASE agency
        -- long live multiline editing
        WHEN	'AG'		THEN	'Air Guard'
        WHEN	'ANC'		THEN	'Alaska Native Corporation'
        WHEN	'BIA'		THEN	'Bureau of Indian Affairs'
        WHEN	'BLM'		THEN	'Bureau of Land Management'
        WHEN	'BOEM'	THEN	'Bureau of Ocean Energy Management'
        WHEN	'BOR'		THEN	'Bureau of Reclamation'
        WHEN	'BSEE'	THEN	'Bureau of Safety and Environmental Enforcement'
        WHEN	'C&L'		THEN	'County & Local'
        WHEN	'CDF'		THEN	'California Department of Forestry & Fire Protection'
        WHEN	'DC'		THEN	'Department of Corrections'
        WHEN	'DFE'		THEN	'Division of Forest Environment'
        WHEN	'DFF'		THEN	'Division of Forestry Fire & State Lands'
        WHEN	'DFL'		THEN	'Division of Forests and Land'
        WHEN	'DFR'		THEN	'Division of Forest Resources'
        WHEN	'DL'		THEN	'Department of Lands'
        WHEN	'DNR'		THEN	'Department of Natural Resources'
        WHEN	'DNRC'	THEN	'Department of Natural Resources and Conservation'
        WHEN	'DNRF'	THEN	'Department of Natural Resources Forest Service'
        WHEN	'DOA'		THEN	'Department of Agriculture'
        WHEN	'DOC'		THEN	'Department of Conservation'
        WHEN	'DOE'		THEN	'Department of Energy'
        WHEN	'DOF'		THEN	'Department of Forestry'
        WHEN	'DVF'		THEN	'Division of Forestry'
        WHEN	'DWF'		THEN	'Division of Wildland Fire'
        WHEN	'EPA'		THEN	'Environmental Protection Agency'
        WHEN	'FC'		THEN	'Forestry Commission'
        WHEN	'FEMA'	THEN	'Federal Emergency Management Agency'
        WHEN	'FFC'		THEN	'Bureau of Forest Fire Control'
        WHEN	'FFP'		THEN	'Forest Fire Protection'
        WHEN	'FFS'		THEN	'Forest Fire Service'
        WHEN	'FR'		THEN	'Forest Rangers'
        WHEN	'FS'		THEN	'Forest Service'
        WHEN	'FWS'		THEN	'Fish & Wildlife Service'
        WHEN	'HQ'		THEN	'Headquarters'
        WHEN	'JC'		THEN	'Job Corps'
        WHEN	'NBC'		THEN	'National Business Center'
        WHEN	'NG'		THEN	'National Guard'
        WHEN	'NNSA'	THEN	'National Nuclear Security Administration'
        WHEN	'NPS'		THEN	'National Park Service'
        WHEN	'NWS'		THEN	'National Weather Service'
        WHEN	'OES'		THEN	'Office of Emergency Services'
        WHEN	'PRI'		THEN	'Private'
        WHEN	'SF'		THEN	'State Forestry'
        WHEN	'SFS'		THEN	'State Forest Service'
        WHEN	'SP'		THEN	'State Parks'
        WHEN	'TNC'		THEN	'The Nature Conservancy'
        WHEN	'USA'		THEN	'United States Army'
        WHEN	'USACE'	THEN	'United States Army Corps of Engineers'
        WHEN	'USAF'	THEN	'United States Air Force'
        WHEN	'USGS'	THEN	'United States Geological Survey'
        WHEN	'USN'		THEN	'United States Navy'
      END AS agency_name
    FROM NWCG_UnitIDActive_20170109
    GROUP BY agency
    ORDER BY 1 DESC")

agencies
data <- data.frame(
  id = agencies$Agency,
  name = agencies$agency_name,
  value = agencies$nr
)
```



```{r fig.width=16}

# https://www.r-graph-gallery.com/296-add-labels-to-circular-barplot
 
# Get the name and the y position of each label
data <- data  %>% filter(name != "<NA>") %>% arrange(value) %>%
   mutate(id=factor(id, id))
label_data <- data %>% filter(name != "<NA>")
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (as.integer(label_data$id)-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)
 
# Make the plot
p <- data %>% filter(name != "<NA>") %>%
  arrange(value) %>%
  mutate(id=factor(id, id)) %>%
  ggplot(aes(x=as.factor(id), y=value)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  geom_bar(stat="identity", fill=alpha("darkgreen", 0.6)) +
  ylim(-100,120) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar(start = 0) + 
  geom_text(data=label_data, aes(x=id, y=value+10, label=name, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=3, angle= label_data$angle, inherit.aes = FALSE ) 
 

p
#ggplot2::ggsave('asd.pdf', p, device = 'pdf')

```


```{r}
# units.GeographicArea is the same as units.gacc, except that California is split into Northern California and Southern California
# as treemap?
total_burnt <- dbGetQuery(
  con,
  "SELECT
     SUM(f.fire_size) AS total,
     CASE units.GeographicArea
        WHEN 'NA' THEN 'National'
        WHEN 'IN' THEN 'International'
        WHEN 'AK' THEN 'Alaska'
        WHEN 'CA' THEN 'California'
        WHEN 'EA' THEN 'Eastern Area'
        WHEN 'GB' THEN 'Great Basin'
        WHEN 'NR' THEN 'Northern Rockies'
        WHEN 'NW' THEN 'Northwest'
        WHEN 'RM' THEN 'Rocky Mountain'
        WHEN 'SA' THEN 'Southern Area'
        WHEN 'SW' THEN 'Southwest'
     END geographic_area
   FROM Fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId GROUP BY units.GeographicArea ORDER BY 1 DESC")

total_burnt

# https://www.r-graph-gallery.com/234-a-very-basic-treemap.html

library(treemap)
# idea: scale by population to determine the most affected regions
treemap::treemap(total_burnt, index = 'geographic_area', vSize = 'total', type = 'index', title = "What regions burned the most?")
```


```{r}
parsed_datetime <- function(col_prefix) {
  glue::glue("datetime(date({col_prefix}_date) || ' ' || substr({col_prefix}_time, 1, 2) || ':' || substr({col_prefix}_time, 3, 4) || ':00')")
}

parsed_time <- function(col_prefix) {
  glue::glue("substr({col_prefix}_time, 1, 2) || ':' || substr({col_prefix}_time, 3, 4) || ':00'")
}
```

```{r}
# containment time - discovery time
all_diffs <- dbGetQuery(con, glue::glue(
  "WITH res AS (
     SELECT
       fire_name,
       fire_size,
       discovery_date,
       cont_date,
       {parsed_datetime('f.discovery')} AS discovery,
       {parsed_datetime('f.cont')} AS contained,
       cont_date - discovery_date AS days_diff,
       time(strftime('%s', {parsed_time('cont')}) - strftime('%s', {parsed_time('discovery')}), 'unixepoch') AS h_diff,
       fire_year,
       fire_size_class
     FROM fires f
     WHERE 
           cont_date IS NOT NULL
       AND cont_time IS NOT NULL
       AND discovery_date IS NOT NULL
       AND discovery_time IS NOT NULL -- AND days_diff > 0
     ORDER BY days_diff DESC, h_diff DESC
     --LIMIT -1 OFFSET 1000
     --LIMIT 40
  )
  
  SELECT days_diff, h_diff, cast(strftime('%s', contained) - strftime('%s', discovery) as real) / (24*3600) AS tm_diff, fire_size_class FROM res", .con=con))
  # SELECT avg(days_diff), fire_size_class FROM res GROUP BY 2", .con=con))
  #SELECT fire_year, fire_size, discovery, contained, days_diff, h_diff FROM res", .con=con))
  #SELECT count(*) from res", .con=con))
#strftime('%s', {parsed_datetime('f.cont')}) - strftime('%s', {parsed_datetime('f.discovery')}) AS duration
#end
all_diffs[1:7,]
```


There seem to be bogus entries: small fires not being estinguished for more than a couple of years:
```{r}
# links: see below

ggplot(all_diffs, aes(x = `days_diff`, y = `fire_size_class`, fill = ..y..)) +
  geom_violin(scale = 3) +
  scale_fill_viridis(name = "days_diff", option = "C") +
  labs(title = 'Days from discovery to containment (raw)') +
  theme_ipsum() +
    theme(
      legend.position="none",
      #panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```

```{r}
# https://www.data-to-viz.com/graph/violin.html
# https://www.data-to-viz.com/caveat/boxplot.html
all_diffs[-8000:0,][1,]

all_diffs2 = all_diffs[-1000:0,] %>% filter(fire_size_class < 'E') %>% filter(tm_diff < 2) # || fire_size_class == 'C')
all_diffs2
sample_size = all_diffs2 %>% group_by(fire_size_class) %>% summarize(num=n())
all_diffs2 %>%
  left_join(sample_size) %>%
  mutate(myaxis = paste0(fire_size_class, "\n", "n=", num)) %>%
ggplot(aes(x = tm_diff, y = myaxis, fill=..y..)) +
  geom_violin(scale = 3) +
    geom_boxplot(width=0.1, color="grey", alpha=0.2) +
  scale_fill_viridis(name = "days_diff", option = "C") +
  labs(title = 'Days from discovery to containment') +
  theme_ipsum() +
    theme(
      legend.position="none",
      #panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```

```{r}
all_data <- dbGetQuery(con, "SELECT DISTINCT (fod_id), * from fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId")
x <- all_data %>% select(STAT_CAUSE_DESCR, DISCOVERY_DOY,UnitType,Department,GeographicArea, OWNER_DESCR, STATE, FIRE_SIZE_CLASS) %>% drop_na() %>% filter(FIRE_SIZE_CLASS < "D")
all_data[1,]
# y <- x$FIRE_SIZE_CLASS
# x <- x %>% select(!matches("FIRE_SIZE_CLASS"))
n <- as.integer(1.4e6)
# nb <- naive_bayes(x = x[1:n,], y = y[1:n], laplace = 1e-5)
df <- sample(as.data.frame(x))
train <- df[1:n,]
n1 <- 100000
test <- df[n+1:n1,]
test2 <- test
test <- test %>% select(!matches("FIRE_SIZE_CLASS"))

nb <- naive_bayes(FIRE_SIZE_CLASS ~ ., train, laplace = 1e-7)
sum(df[n+1:n1,]$FIRE_SIZE_CLASS == predict(nb, test, type = "class"))/n1
preds <- predict(nb, test, type = "class")
test2$pred <- preds
#as.tibble(df[n+1:n1,]$FIRE_SIZE_CLASS) %>% conf_mat(preds, fire_size_class)
test2 %>% conf_mat(pred, FIRE_SIZE_CLASS)
```

```{r}
dbDisconnect(conn = con)
```


```{r}
# misc
# units not found in units table
# dbGetQuery(con, "with ids as (SELECT unitId from NWCG_UnitIDActive_20170109) select count(distinct NWCG_REPORTING_UNIT_ID) from fires where NWCG_REPORTING_UNIT_ID not in ids")
#   count(distinct NWCG_REPORTING_UNIT_ID)
# 1                                     17
# 34757 fire records belong to these units
```

```{r}

# ??
#dbGetQuery(con, "SELECT COUNT(*), WildlandRole FROM NWCG_UnitIDActive_20170109 GROUP BY 2")
#   COUNT(*)                 WildlandRole
# 1      328 Dispatch/Coordination Center
# 2       29            Fire/Radio Cashes
# 3        8     Incident Host Functional
# 4     5228     Incident Host Geographic
# 5      266       Resource Provider Only
# 6        8            Training Facility

# > dbGetQuery(con, "SELECT COUNT(*) FROM NWCG_UnitIDActive_20170109 WHERE Parent IS NULL")
#   COUNT(*)
# 1     5867
# > dbGetQuery(con, "SELECT COUNT(*) FROM NWCG_UnitIDActive_20170109 WHERE Parent IS NOT NULL")
#   COUNT(*)
# 1        0
```
