---
title: "General EDA plots and models"
output:
  html_notebook:
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: true
---

[Homepage](https://mcmarius.github.io/InformationVisualization/)

```{r}
library(RSQLite)
library(tidyverse)
library(hrbrthemes)
library(kableExtra)
library(RColorBrewer)
library(viridis)
library(d3heatmap)
library(naivebayes)
library(yardstick)
library(ranger)
library(caret)
library(treemap)
# library(d3treeR)
library(highcharter)
#library(Metrics)
# library(ggridges)
```

```{r}
con <- dbConnect(RSQLite::SQLite(), "../wildfire/FPA_FOD_20170508.sqlite")
causes <- dbGetQuery(con, "SELECT COUNT(*) AS nr, STAT_CAUSE_DESCR AS cause_name FROM Fires GROUP BY STAT_CAUSE_DESCR")

#knitr::opts_chunk$set(fig.width=16)

# https://www.reddit.com/r/RStudio/comments/euew3l/preview_button_gone_can_only_knit/

# some color ideas from https://colorbrewer2.org
```


# Main causes

## What are the causes for wildfires?

```{r}
# https://www.data-to-viz.com/graph/lollipop.html

causes %>%
  arrange(nr) %>%
  mutate(cause_name=factor(cause_name, cause_name)) %>%
  ggplot( aes(x=cause_name, y=nr) ) +
    geom_segment(aes(x=cause_name, xend=cause_name, y=0, yend=nr), color="darkgrey") +
    geom_point(size=4, color='#e34a33', alpha=0.9) +
    coord_flip() +
    theme_ipsum() +
    # xlab("") +
    labs(x="Cause",y="Number of fires", title = 'Main causes for wildfires') +
    #labs(x="",y="", title = 'Main causes for wildfires') +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      #legend.position="none",
      #axis.text = element_text(hjust = 0.5)
    )
```

```{r}
fire_size_class <- dbGetQuery(con, "SELECT COUNT(*) AS nr, FIRE_SIZE_CLASS AS fire_size FROM Fires GROUP BY FIRE_SIZE_CLASS")
```

## Almost 2 million fires? Is there anything left?

```{r}
# https://www.data-to-viz.com/graph/barplot.html

fire_size_class %>%
  arrange(nr) %>%
  mutate(fire_size=factor(fire_size, fire_size)) %>%
  ggplot( aes(x=fire_size, y=nr) ) +
    #geom_segment(aes(x=cause_name, xend=fire_size, y=0, yend=nr), color="darkgrey") +
    #geom_point(size=4, color='#e34a33', alpha=0.9) +
    geom_bar(stat='identity', fill='darkred', width = 0.75) +
    coord_flip() +
    #theme_ipsum() +
    #xlab("") +
    #labs(x="",y="", title = 'Number of fires by size') +
    labs(x="Fire size class",y="Number of fires", title = 'Number of fires by size') +
    theme(
      panel.grid.minor.x = element_line(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      #axis.ticks.x = fire_size_class$nr
      #legend.position="none",
      #axis.text = element_text(hjust = 0.5)
    ) # + scale_y_discrete(fire_size_class$nr)
```

## What are the most frequent causes for the biggest fires?

Note that these are normalized by column (fire class).

```{r}
# https://www.data-to-viz.com/graph/heatmap.html
# https://www.r-graph-gallery.com/215-the-heatmap-function.html

fire_cause <- dbGetQuery(con,
                         "SELECT STAT_CAUSE_DESCR AS cause, FIRE_SIZE_CLASS AS fire_class, COUNT(FIRE_SIZE_CLASS) AS nr
                         FROM Fires
                         WHERE FIRE_SIZE_CLASS IN ('D', 'E', 'F', 'G')
                         GROUP BY 1, 2")
```

```{r}

nrows <- length(levels(factor(fire_cause$cause)))
ncols <- length(levels(factor(fire_cause$fire_class)))
matr3 <- t(matrix(fire_cause$nr, nrow = ncols, ncol = nrows))
rownames(matr3) <- levels(factor(fire_cause$cause))
colnames(matr3) <- levels(factor(fire_cause$fire_class))
coul <- colorRampPalette(brewer.pal(8, "OrRd"))(25)
d3heatmap(matr3, scale = 'column', Colv = NA, Rowv = NA, col=coul,  yaxis_width = 200, width = 700, title = 'The most frequent causes for the biggest fires')
```



```{r}
# sources: same as previous heatmap

sfire_cause <- dbGetQuery(con,
                         "SELECT STAT_CAUSE_DESCR AS cause, FIRE_SIZE_CLASS AS fire_class, COUNT(FIRE_SIZE_CLASS) AS nr
                         FROM Fires
                         WHERE FIRE_SIZE_CLASS IN ('A', 'B', 'C')
                         GROUP BY 1, 2")

```

```{r}
snrows <- length(levels(factor(sfire_cause$cause)))
sncols <- length(levels(factor(sfire_cause$fire_class)))
smatr3 <- t(matrix(sfire_cause$nr, nrow = sncols, ncol = snrows))
rownames(smatr3) <- levels(factor(sfire_cause$cause))
colnames(smatr3) <- levels(factor(sfire_cause$fire_class))
scoul <- colorRampPalette(brewer.pal(8, "OrRd"))(25)
d3heatmap(smatr3, scale = 'column', Colv = NA, Rowv = NA, col=scoul, yaxis_width = 200, width = 700, main = 'The most frequent causes for small fires')
```



```{r}
all_fire_cause <- dbGetQuery(con,
                         "SELECT STAT_CAUSE_DESCR AS cause, FIRE_SIZE_CLASS AS fire_class, COUNT(FIRE_SIZE_CLASS) AS nr
                         FROM Fires
                         GROUP BY 1, 2")
anrows <- length(levels(factor(all_fire_cause$cause)))
ancols <- length(levels(factor(all_fire_cause$fire_class)))
amatr3 <- t(matrix(all_fire_cause$nr, nrow = ancols, ncol = anrows))
rownames(amatr3) <- levels(factor(all_fire_cause$cause))
colnames(amatr3) <- levels(factor(all_fire_cause$fire_class))
acoul <- colorRampPalette(brewer.pal(8, "OrRd"))(25)
d3heatmap(amatr3, scale = 'column', Colv = NA, Rowv = NA, col=acoul, yaxis_width = 200, width = 700, main = 'The most frequent causes for all fires')
```

# Time series
## Are there fires all year long?

[Let us take a closer look through the lens of time.](Timeseries)


```{r}
# source: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

# end of multiplot
# source: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
##############################################################################

```

```{r}

units_by_type <- dbGetQuery(con, "SELECT count(distinct(units.UnitId)) AS nr_units, count(distinct(f.fod_id)) AS nr_fires, units.UnitType FROM Fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId GROUP BY units.UnitType")

# units_by_type
# nr_units | nr_fires | UnitType
#     7      	   40    	Interagency		
#    11        3725   	Tribe		
#   526       86871   	US County/Local		
#   902   	 477787 	  US Federal		
#   177	    1277285   	US State
```

# Another possible explanation besides global warming
## How many wildfires are managed by each unit type?

The following pattern might uncover organizational issues.


```{r}

# https://www.r-graph-gallery.com/37-barplot-with-number-of-observation.html
data <- data.frame(
  counts = units_by_type[,1:2],#) %>% map_int(1L)),
  units = rep(units_by_type[,3], times=1)
)

# same as the barplot above
# colors picked to show fires with some sort of red and units with some sort of blue (obvious reasons)

p1 <- data  %>% 
  #mutate(year=as.factor(year)) %>%
  ggplot( aes(x=units, y=counts.nr_fires)) +
    geom_bar(stat="identity", fill='#e6550d') +
    scale_fill_viridis(discrete=TRUE, name="") +
    theme_ipsum() +
    xlab("Unit type")+
    ylab("Number of fires")

p2 <- data  %>% 
  #mutate(year=as.factor(year)) %>%
  ggplot( aes(x=units, y=counts.nr_units)) +
    geom_bar(stat="identity", fill='#2b8cbe') +
    scale_fill_viridis(discrete=TRUE, name="") +
    theme_ipsum() +
    xlab("Unit type")+
    ylab("Number of units")
multiplot(p1, p2)

```


Further down the line, we notice many agencies each with very few units.

```{r}
# units per agency
# units names taken from kaggle dataset description

agencies <- dbGetQuery(
  con, "
    SELECT
      COUNT(*) AS nr,
      --unitType,
      agency,
      CASE agency
        -- long live multiline editing
        WHEN	'AG'		THEN	'Air Guard'
        WHEN	'ANC'		THEN	'Alaska Native Corporation'
        WHEN	'BIA'		THEN	'Bureau of Indian Affairs'
        WHEN	'BLM'		THEN	'Bureau of Land Management'
        WHEN	'BOEM'	THEN	'Bureau of Ocean Energy Management'
        WHEN	'BOR'		THEN	'Bureau of Reclamation'
        WHEN	'BSEE'	THEN	'Bureau of Safety and Environmental Enforcement'
        WHEN	'C&L'		THEN	'County & Local'
        WHEN	'CDF'		THEN	'California Department of Forestry & Fire Protection'
        WHEN	'DC'		THEN	'Department of Corrections'
        WHEN	'DFE'		THEN	'Division of Forest Environment'
        WHEN	'DFF'		THEN	'Division of Forestry Fire & State Lands'
        WHEN	'DFL'		THEN	'Division of Forests and Land'
        WHEN	'DFR'		THEN	'Division of Forest Resources'
        WHEN	'DL'		THEN	'Department of Lands'
        WHEN	'DNR'		THEN	'Department of Natural Resources'
        WHEN	'DNRC'	THEN	'Department of Natural Resources and Conservation'
        WHEN	'DNRF'	THEN	'Department of Natural Resources Forest Service'
        WHEN	'DOA'		THEN	'Department of Agriculture'
        WHEN	'DOC'		THEN	'Department of Conservation'
        WHEN	'DOE'		THEN	'Department of Energy'
        WHEN	'DOF'		THEN	'Department of Forestry'
        WHEN	'DVF'		THEN	'Division of Forestry'
        WHEN	'DWF'		THEN	'Division of Wildland Fire'
        WHEN	'EPA'		THEN	'Environmental Protection Agency'
        WHEN	'FC'		THEN	'Forestry Commission'
        WHEN	'FEMA'	THEN	'Federal Emergency Management Agency'
        WHEN	'FFC'		THEN	'Bureau of Forest Fire Control'
        WHEN	'FFP'		THEN	'Forest Fire Protection'
        WHEN	'FFS'		THEN	'Forest Fire Service'
        WHEN	'FR'		THEN	'Forest Rangers'
        WHEN	'FS'		THEN	'Forest Service'
        WHEN	'FWS'		THEN	'Fish & Wildlife Service'
        WHEN	'HQ'		THEN	'Headquarters'
        WHEN	'JC'		THEN	'Job Corps'
        WHEN	'NBC'		THEN	'National Business Center'
        WHEN	'NG'		THEN	'National Guard'
        WHEN	'NNSA'	THEN	'National Nuclear Security Administration'
        WHEN	'NPS'		THEN	'National Park Service'
        WHEN	'NWS'		THEN	'National Weather Service'
        WHEN	'OES'		THEN	'Office of Emergency Services'
        WHEN	'PRI'		THEN	'Private'
        WHEN	'SF'		THEN	'State Forestry'
        WHEN	'SFS'		THEN	'State Forest Service'
        WHEN	'SP'		THEN	'State Parks'
        WHEN	'TNC'		THEN	'The Nature Conservancy'
        WHEN	'USA'		THEN	'United States Army'
        WHEN	'USACE'	THEN	'United States Army Corps of Engineers'
        WHEN	'USAF'	THEN	'United States Air Force'
        WHEN	'USGS'	THEN	'United States Geological Survey'
        WHEN	'USN'		THEN	'United States Navy'
      END AS agency_name
    FROM NWCG_UnitIDActive_20170109
    WHERE agency IS NOT NULL
    GROUP BY agency --, unitType
    ORDER BY 1 DESC")

agencies
# 398 units with NULL agency
data <- data.frame(
  id = agencies$Agency,
  name = agencies$agency_name,
  value = agencies$nr
)
```


## There are many agencies with very few units each

```{r fig.width=16}

# https://www.r-graph-gallery.com/296-add-labels-to-circular-barplot
 
# Get the name and the y position of each label
data <- data  %>% filter(name != "<NA>") %>% arrange(value) %>%
   mutate(id=factor(id, id))
label_data <- data %>% filter(name != "<NA>")
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (as.integer(label_data$id)-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)
 
# Make the plot
p <- data %>% filter(name != "<NA>") %>%
  arrange(value) %>%
  mutate(id=factor(id, id)) %>%
  ggplot(aes(x=as.factor(id), y=value)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  geom_bar(stat="identity", fill=alpha("darkgreen", 0.6)) +
  ylim(-100,120) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar(start = 0) + 
  #labs(title = 'Number of units per agency') +
  geom_text(data=label_data, aes(x=id, y=value+10, label=name, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=3, angle= label_data$angle, inherit.aes = FALSE ) 
 

p
#ggplot2::ggsave('circular_plot.pdf', p, device = 'pdf')

```

So indeed it might also be an administrative problem, as the previous plots suggest. Despite having many agencies, these agencies do not have enough units.

```{r}

agencies_by_unit_type <- dbGetQuery(con, "SELECT count(distinct(agency)) AS nr_agencies, units.UnitType FROM NWCG_UnitIDActive_20170109 units GROUP BY units.UnitType")
agencies_by_unit_type
```


```{r}
state_agencies <- dbGetQuery(
  con, "
    SELECT
      COUNT(*) AS nr,
      agency    
    FROM NWCG_UnitIDActive_20170109
    WHERE unitType = 'US State'
    GROUP BY agency
    ORDER BY 1 DESC")
state_agencies
```


## Noisy data

```{r}
parsed_datetime <- function(col_prefix) {
  glue::glue("datetime(date({col_prefix}_date) || ' ' || substr({col_prefix}_time, 1, 2) || ':' || substr({col_prefix}_time, 3, 4) || ':00')")
}

parsed_time <- function(col_prefix) {
  glue::glue("substr({col_prefix}_time, 1, 2) || ':' || substr({col_prefix}_time, 3, 4) || ':00'")
}
```

```{r}
# containment time - discovery time
all_diffs <- dbGetQuery(con, glue::glue(
  "WITH res AS (
     SELECT
       fire_name,
       fire_size,
       discovery_date,
       cont_date,
       {parsed_datetime('f.discovery')} AS discovery,
       {parsed_datetime('f.cont')} AS contained,
       cont_date - discovery_date AS days_diff,
       time(strftime('%s', {parsed_time('cont')}) - strftime('%s', {parsed_time('discovery')}), 'unixepoch') AS h_diff,
       fire_year,
       fire_size_class
     FROM fires f
     WHERE 
           cont_date IS NOT NULL
       AND cont_time IS NOT NULL
       AND discovery_date IS NOT NULL
       AND discovery_time IS NOT NULL -- AND days_diff > 0
     ORDER BY days_diff DESC, h_diff DESC
     --LIMIT -1 OFFSET 1000
     --LIMIT 40
  )
  
  SELECT days_diff, h_diff, cast(strftime('%s', contained) - strftime('%s', discovery) as real) / (24*3600) AS tm_diff, fire_size_class FROM res", .con=con))
  # SELECT avg(days_diff), fire_size_class FROM res GROUP BY 2", .con=con))
  #SELECT fire_year, fire_size, discovery, contained, days_diff, h_diff FROM res", .con=con))
  #SELECT count(*) from res", .con=con))
#strftime('%s', {parsed_datetime('f.cont')}) - strftime('%s', {parsed_datetime('f.discovery')}) AS duration
#end
all_diffs[1:7,]
```


There seem to be bogus entries: small fires not being estinguished for more than a couple of years

```{r}
# links: see below

ggplot(all_diffs, aes(x = `days_diff`, y = `fire_size_class`, fill = ..y..)) +
  geom_violin(scale = 3) +
  scale_fill_viridis(name = "days_diff", option = "C") +
  labs(title = 'Days from discovery to containment (raw)', x = "Days", y = "Fire size class") +
  theme_ipsum() +
    theme(
      legend.position="none",
      #panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```

We remove the first thousand entries and fires lasting more than 2 days

```{r}
# https://www.data-to-viz.com/graph/violin.html
# https://www.data-to-viz.com/caveat/boxplot.html
all_diffs[-8000:0,][1,]

all_diffs2 = all_diffs[-1000:0,] %>% filter(fire_size_class < 'E') %>% filter(tm_diff < 2) # || fire_size_class == 'C')
all_diffs2
sample_size = all_diffs2 %>% group_by(fire_size_class) %>% summarize(num=n())
all_diffs2 %>%
  left_join(sample_size) %>%
  mutate(fire_size = paste0(fire_size_class, "\n", "n=", num)) %>%
ggplot(aes(x = tm_diff, y = fire_size, fill=..y..)) +
  geom_violin(scale = 3) +
    geom_boxplot(width=0.1, color="grey", alpha=0.2) +
  scale_fill_viridis(name = "days_diff", option = "C") +
  labs(title = 'Days from discovery to containment (small fires)', x = "Days", y = "Fire size class") +
  theme_ipsum() +
    theme(
      legend.position="none",
      #panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```

We remove fires lasting more than a year

```{r}
all_diffs3 = all_diffs %>% filter(fire_size_class > 'D') %>% filter(tm_diff < 365) # || fire_size_class == 'C')
all_diffs3
sample_size = all_diffs3 %>% group_by(fire_size_class) %>% summarize(num=n())
all_diffs3 %>%
  left_join(sample_size) %>%
  mutate(fire_size = paste0(fire_size_class, "\n", "n=", num)) %>%
ggplot(aes(x = tm_diff, y = fire_size, fill=..y..)) +
  geom_violin(scale = 3) +
    geom_boxplot(width=0.1, color="grey", alpha=0.2) +
  scale_fill_viridis(name = "days_diff", option = "C") +
  labs(title = 'Days from discovery to containment (large fires)', x = "Days", y = "Fire size class") +
  theme_ipsum() +
    theme(
      legend.position="none",
      #panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```




# Geographic data
## What regions burned the most?

```{r}
# units.GeographicArea is the same as units.gacc, except that California is split into Northern California and Southern California
# as treemap
total_burnt <- dbGetQuery(
  con,
  "SELECT
     SUM(f.fire_size) AS total,
     CASE units.GeographicArea
        WHEN 'NA' THEN 'National'
        WHEN 'IN' THEN 'International'
        WHEN 'AK' THEN 'Alaska'
        WHEN 'CA' THEN 'California'
        WHEN 'EA' THEN 'Eastern Area'
        WHEN 'GB' THEN 'Great Basin'
        WHEN 'NR' THEN 'Northern Rockies'
        WHEN 'NW' THEN 'Northwest'
        WHEN 'RM' THEN 'Rocky Mountain'
        WHEN 'SA' THEN 'Southern Area'
        WHEN 'SW' THEN 'Southwest'
     END geographic_area
   FROM Fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId GROUP BY units.GeographicArea ORDER BY 1 DESC")

total_burnt
```

```{r}
# https://www.r-graph-gallery.com/234-a-very-basic-treemap.html

treemap::treemap(total_burnt, index = 'geographic_area', vSize = 'total', type = 'index', title = "What regions burned the most?")
```


```{r}
# https://github.com/jakevdp/data-USstates
abbrevs <- read.csv('us_state-abbrevs.csv')
census <- read.csv('us_state-population.csv') %>%
  filter(ages == "total" & year == 2012) %>%
  merge(abbrevs, by.x = c("state.region"), by.y = c("abbreviation")) %>%
  select(state, population)

# note: Alaska and California regions have been renamed because the highcharts
# library requires unique labels across all levels
total_burnt2 <- dbGetQuery(
  con,
  "SELECT
     SUM(f.fire_size) AS total, f.state AS abbreviation,
     CASE units.GeographicArea
        WHEN 'NA' THEN 'National'
        WHEN 'IN' THEN 'International'
        WHEN 'AK' THEN 'Alaska (region)'
        WHEN 'CA' THEN 'California (region)'
        WHEN 'EA' THEN 'Eastern Area'
        WHEN 'GB' THEN 'Great Basin'
        WHEN 'NR' THEN 'Northern Rockies'
        WHEN 'NW' THEN 'Northwest'
        WHEN 'RM' THEN 'Rocky Mountain'
        WHEN 'SA' THEN 'Southern Area'
        WHEN 'SW' THEN 'Southwest'
     END geographic_area
   FROM Fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId
   GROUP BY units.GeographicArea, f.state
   ORDER BY 1 DESC") %>%
# total_burnt2 <-
  # dbGetQuery(con, "SELECT SUM(f.fire_size) AS total, state,  FROM Fires f GROUP BY state ORDER BY 1 DESC") %>%
  merge(abbrevs) %>% #, by.x = c("state"), by.y = c("abbreviation")) %>%
  merge(census)
total_burnt2$affected <- (total_burnt2$population * total_burnt2$total)
```

**Important notice!** A state can be in one or more regions. This is how regions are defined in the Fires dataset. This is an issue for our treemaps, because we have no means of "splitting" a state population by region. This issue is only strikingly visible with California, so we manually force it to be just in California region.

We still have duplicate entries, but they should not affect the overall visualization.

```{r}
# https://www.datanovia.com/en/lessons/highchart-interactive-treemap-in-r/
# https://www.rdocumentation.org/packages/highcharter/versions/0.8.2/topics/hctreemap
# https://api.highcharts.com/highcharts/plotOptions.treemap.allowTraversingTree
# options(highcharter.theme = hc_theme_smpl(tooltip = list(valueDecimals = 2)))

total_burnt2
# tree <- treemap(
#   total_burnt2,
#   index = c('geographic_area', 'state'),
#   vSize = 'affected',
#   type = 'index',
#   title = "What states were the most affected?",
#   align.labels = list(c('center', 'center'), c('center', 'center')))
total_burnt22 <- total_burnt2 %>% filter((geographic_area != "California (region)" & state != "California") | geographic_area == "California (region)")
```


```{r}
hchart(
  data_to_hierarchical(total_burnt22, group_vars = c(geographic_area, state), size_var = 'population'),
  type = "treemap",
  allowTraversingTree = T,
  layoutAlgorithm = "sliceAndDice",  # squarified
  layoutStartingDirection = "vertical"  # horizontal
) %>% hc_title(text = 'US regions by population')
```

## The most affected regions

```{r}
hchart(
  data_to_hierarchical(total_burnt2, group_vars = c(geographic_area, state), size_var = 'affected'),
  type = "treemap", allowTraversingTree = T,
  layoutAlgorithm = "sliceAndDice",  # squarified
  layoutStartingDirection = "vertical"  # horizontal
) %>%
  hc_title(text = 'US regions by burned area and population')
  #hc_tooltip(enabled = F)
# d3tree(tree, 'What regions and states were the most affected?', width = '780', height = '450')
```


## Deeper dive into geographic EDA

The story continues [here](geo_eda_plots) and [here](models).


# Are fires predictable?

Unfortunately, data for large fires is incomplete or missing.

Therefore, the following prediction tasks are limited to small fires.

## Naive Bayes to predict the size of a fire

```{r}
# https://yardstick.tidymodels.org/reference/conf_mat.html
all_data <- dbGetQuery(con, "SELECT DISTINCT (fod_id), * from fires f INNER JOIN NWCG_UnitIDActive_20170109 units ON f.NWCG_REPORTING_UNIT_ID = units.UnitId")
x <- all_data %>% select(STAT_CAUSE_DESCR, DISCOVERY_DOY,UnitType,Department,GeographicArea, OWNER_DESCR, STATE, FIRE_SIZE_CLASS) %>% drop_na() %>% filter(FIRE_SIZE_CLASS < "D")

all_data[1,]

# nb <- naive_bayes(x = x[1:n,], y = y[1:n], laplace = 1e-5)
df <- sample(as.data.frame(x))

data_split <- df$FIRE_SIZE_CLASS %>% createDataPartition(p = 0.8, list = FALSE)
train  <- df[data_split, ]
test <- df[-data_split, ]

test2 <- test
test <- test %>% select(!matches("FIRE_SIZE_CLASS"))

nb <- naive_bayes(FIRE_SIZE_CLASS ~ ., train, laplace = 1e-7)
preds <- predict(nb, test, type = "class")
# accuracy
sum(test2$FIRE_SIZE_CLASS == preds)/length(test2$FIRE_SIZE_CLASS)
# 0.5924989
test2$pred <- preds
#as.tibble(df[n+1:n1,]$FIRE_SIZE_CLASS) %>% conf_mat(preds, fire_size_class)
```


```{r}
cm <- test2 %>% conf_mat(pred, FIRE_SIZE_CLASS)

autoplot(cm, type="heatmap")+scale_fill_gradient(low="white", high="#009194")
```

## Random forest to predict the cause of a fire

```{r}
# https://stackoverflow.com/a/64539733
# https://www.datanovia.com/en/blog/ggplot-axis-ticks-set-and-rotate-text-labels/
# https://bradleyboehmke.github.io/HOML/random-forest.html

x <- all_data %>% select(STAT_CAUSE_DESCR, DISCOVERY_DOY,UnitType,Department,GeographicArea, OWNER_DESCR, STATE, FIRE_SIZE, FIRE_SIZE_CLASS, CONT_DOY, SOURCE_SYSTEM_TYPE, NWCG_REPORTING_UNIT_NAME, LATITUDE, LONGITUDE, SOURCE_REPORTING_UNIT_NAME) %>% drop_na() %>% filter(FIRE_SIZE_CLASS < "D") %>% select(!matches("FIRE_SIZE_CLASS"))
x$STAT_CAUSE_DESCR <- as.factor(x$STAT_CAUSE_DESCR)

df <- sample(as.data.frame(x))
data_split <- df$STAT_CAUSE_DESCR %>% createDataPartition(p = 0.8, list = FALSE)
train  <- df[data_split, ]
test <- df[-data_split, ]

test2 <- test
test <- test %>% select(!matches("STAT_CAUSE_DESCR"))

model3 <- ranger(STAT_CAUSE_DESCR ~ ., train, num.trees = 20)

model3$confusion.matrix

#test2$STAT_CAUSE_DESCR <- as.factor(test2$STAT_CAUSE_DESCR)
preds <- predict(model3, data=test)
test2$preds <- preds$predictions

# accuracy
sum(test2$STAT_CAUSE_DESCR == test2$preds)/length(test2$preds)
# [1] 0.5819649  # 0.5835164
# baseline - majority class
sum(test2$STAT_CAUSE_DESCR == "Lightning")/length(test2$STAT_CAUSE_DESCR)
#[1] 0.2286784
```

```{r}
cm <- test2 %>% conf_mat(preds, STAT_CAUSE_DESCR)
autoplot(cm, type="heatmap")+scale_fill_gradient(low="white", high="#009194") +theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#dbDisconnect(conn = con)
```


```{r}
# misc
# units not found in units table
# dbGetQuery(con, "with ids as (SELECT unitId from NWCG_UnitIDActive_20170109) select count(distinct NWCG_REPORTING_UNIT_ID) from fires where NWCG_REPORTING_UNIT_ID not in ids")
#   count(distinct NWCG_REPORTING_UNIT_ID)
# 1                                     17
# 34757 fire records belong to these units
```

```{r}

# ??
#dbGetQuery(con, "SELECT COUNT(*), WildlandRole FROM NWCG_UnitIDActive_20170109 GROUP BY 2")
#   COUNT(*)                 WildlandRole
# 1      328 Dispatch/Coordination Center
# 2       29            Fire/Radio Cashes
# 3        8     Incident Host Functional
# 4     5228     Incident Host Geographic
# 5      266       Resource Provider Only
# 6        8            Training Facility

# > dbGetQuery(con, "SELECT COUNT(*) FROM NWCG_UnitIDActive_20170109 WHERE Parent IS NULL")
#   COUNT(*)
# 1     5867
# > dbGetQuery(con, "SELECT COUNT(*) FROM NWCG_UnitIDActive_20170109 WHERE Parent IS NOT NULL")
#   COUNT(*)
# 1        0
```
